Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.tex
Until diagnostics

\begin{abstract}
Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence of iterative imputation algorithms. We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
\end{abstract}

\keywords{missing data, iterative imputation, non-convergence, mice}

@@

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous across science, but treating it can also be a tedious task. If a dataset contains just one incomplete observation, calculations are not defined, and statistical models cannot be fitted to the data. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this \emph{ad hoc} solution may yield invalid results \citep{buur18}. An alternative is to apply imputation. With imputation, we `fill in' the missing values in an incomplete dataset. Subsequently, the model of scientific interest can be fitted to the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances \citep{buur18}.

@@

\caption{Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
\end{figure}

A popular method to obtain imputations \(\dot y_{\rm imp, \ell}\) is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values \(y_{\rm mis}\). The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to formulate an informed advice on when approxiamte convergence may be safely concluded. We will therefore assess the performance of different methods to identify non-convergence. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}. In iterative imputation algorithms mixing implies that imputation chains intermingle nicely, and stationarity is characterized by \textbf{the absence of trending} within chains. If one of these two requirements is not met, there is non-convergence in the algorithm. Without mixing, chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. The distribution of imputed values then differs across imputations. This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a systematically lower or higher set of imputations. Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.

To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\). The \(\theta\) in this example is the chain mean of a variable that we will call \(j\) (i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\)). In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. Non-convergence thus influences the mean of \(j\) in \(\dot y_{\rm imp, \ell}\). This difference (presumably) translates to the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently to the pooled estimate \(\bar{\mu}_j\). Ergo, non-convergence leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}MERGE with sections below{]}} Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number. Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation. Monitoring convergence by inspecting traceplots of these \(\theta\)s may be undesirable for several reasons: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} and, 3) there is not an objective measure that quantifies convergence.

\textbf{{[}MERGE with section above and below{]}} Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of \(y_{\rm imp}\). Ideally, a multivariate \(\theta\) should be monitored. A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature. Implementing this, however, might be somewhat too technical for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. On these grounds, such a \(\theta\) can be considered insufficient too. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.

\textbf{{[}MERGE with sections above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The default scalar summaries in \texttt{mice} are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_{1, \ell}\) as such a scalar summary. We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets. The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.\textbf{{[}ADD definition variance in first PCA component is equal to the first eigenvalue of the variance-covariance matrix{]}}

>>>>

\begin{abstract}
Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence in iterative imputation algorithms. We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
\end{abstract}

\keywords{missing data, iterative imputation, non-convergence, MICE}

@@

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous across science, but treating it can also be a tedious task. If a dataset contains just one incomplete observation, calculations are not defined, and statistical models cannot be fitted to the data. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this \emph{ad hoc} solution may yield invalid results \citep{buur18}. An alternative is to apply imputation. With imputation, we `fill in' the missing values in an incomplete dataset. Subsequently, the model of scientific interest can be fitted to the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances \citep{buur18}.

@@

\caption{Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\hat{Q}$, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
\end{figure}

A popular method to obtain imputations is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values. The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, \textbf{methods} may only identify signs of \emph{non}-convergence \citep{hoff09}. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to formulate an informed advice on when to safely conclude approximate convergence. We will therefore assess the performance of different methods to identify non-convergence. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}. In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by \textbf{the absence of trending} within chains. If one of these two requirements is not met, we speak of non-convergence. Without mixing, chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. The distribution of imputed values then differs across imputations. This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a systematically lower or higher set of imputations. Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.

\hypertarget{non-convergence-in-mice}{%
\subsection{Non-convergence in MICE}\label{non-convergence-in-mice}}

To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\). The \(\theta\) in this example is the average of the imputed values for variable \(j\) in each imputation \(\dot y_{\rm imp, \ell}\) (also called the `chain mean' of \(j\)). In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence impacts the distribution of the imputed values in this example. This effect may translate into the distribution of the completed data, which consequently affects the estimated quantity of scientific interest \(\bar{Q}\). \(\bar{Q}\) may be biased due to systematically higher or lower imputed values, and invalid because the variance within imputation chains is under-estimated. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms. \textbf{{[}ADD t and l to each theta??{]}}

@@

\hypertarget{identifying-non-convergence}{%
\subsection{Identifying non-convergence}\label{identifying-non-convergence}}

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}. In \texttt{mice}, the default scalar summaries to inspect are chain means and chain variances.

Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}. Moreover, traceplots typically only display chain means and chain variances, which are univariate scalar summaries of the state space of the algorithm. Monitoring univariate \(\theta\)s may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of \(\dot y_{\rm imp, \ell}\). Ideally, a multivariate \(\theta\) should be monitored.

A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature. Implementing this, however, might be somewhat too technical for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. Additionally, one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently. On these grounds, such a \(\theta\) can be considered insufficient too.

As alternative, van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.

\hypertarget{a-novel-scalar-summary}{%
\subsection{A novel scalar summary}\label{a-novel-scalar-summary}}

The downside to the current \(\theta\)s to evaluate is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent \(\theta\) that summarizes the multivariate nature of the data.

We propose \(\lambda_{1, \ell}\) as such a scalar summary. We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets. The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.\textbf{{[}ADD definition variance in first PCA component is equal to the first eigenvalue of the variance-covariance matrix{]}}
